---
title: "Prediction Assignment Writeup"
author: "MF"
date: "19 November 2017"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
# Summary
Using tracking devices such as Jawbone Up, Nike FuelBand, and Fitbit people who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants will be used to classify how well the exercise has been performed. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. Details on the data can be found here:  http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (Weight Lifting Exercise Dataset).

The training data set used was:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data set used was:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The test data provided was only used after cross validating the model on the training data. This was done by splitting the training data into new training and testing data sets

The model has a 98.5% accuracy on the new training set, 99% accuarcy on the new testing data set and 100% corrrect for the Quiz data set(quiz answers checked externally).

##Modelling 
###Downloading the Data
```{r data, cache = TRUE, eval = TRUE,echo=TRUE}
download.file('https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv','training.csv')
traindata<-read.csv('training.csv', na.strings=c("NA","#DIV/0!",""))
traindata<-as.data.frame(traindata)
```
```{r dataTEST, cache = TRUE, eval = TRUE,echo=TRUE}
download.file('https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv','test.csv')
testdata<-read.csv('test.csv', na.strings=c("NA","#DIV/0!",""))
testdata<-as.data.frame(testdata)
```
###Data Exploration and Preprocessing
The data has 160 variables which is a large amount to try use in a model. Therefore after looking at the data, 3 things were noticed:
1. The first 7 columns are variables that will not affect the outcome
2. There are a lot of variables that do no vary significantly
3. There are a lot of variables which mostly have NA values  

To deal with these 3 features of the data, the data was processed in 3 ways:
1. The first 7 columns were removed
2. Near Zero Variance analysis was used to remove 101 variables
3. any columns still with NA values were removed.

The training data was then split into a smaller training data set and a new testing set to cross validate the model before trying it on the quiz data provided.

```{r preprocess, cache = TRUE, eval =TRUE ,echo=TRUE}
library(caret)
set.seed(33221)
dim((traindata))
##Remove top first 7, which are not valid variables.
c<-c(1:7)
traindata1<-traindata[,-c]
testdata1<-testdata[,-c]
##checking for near zero variability in the variables
nsv<-nearZeroVar(testdata,freqCut=2,uniqueCut=20,saveMetrics = F)
traindata2<-traindata1[,-nsv]
testdata2<-testdata1[,-nsv]
##checking for columns with NA values and removing those columns from the variables
traindata3<-traindata2[,names(traindata2[,colSums(is.na(traindata2)) == 0])]
names(testdata2)[names(testdata2) == "problem_id"] <- "classe"
testdata3<-testdata2[,names(traindata2[,colSums(is.na(traindata2)) == 0])]
##Partitioning the training set to do cross validation
inTrain <- createDataPartition(traindata3$classe, p=0.6, list=FALSE)
training <- traindata3[inTrain,]
testing <- traindata3[-inTrain,]
dim(training); dim(testing);
```
Above are the dimensions of the newly created training and testing set.
###Creating the Model
The first model fitted oto the data was Random Forest. This was tried first as it was the method used in by Velloso et al., 2013 in the original study.
```{r process, cache = TRUE, eval = TRUE,echo=TRUE}
library(parallel)
library(doParallel)
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)
fitControl <- trainControl(method = "cv",number = 5, allowParallel = TRUE)    
model<-train(classe~.,method="rf",data=training,trcontrol=fitControl)
```
```{r model, cache = F, eval = TRUE,echo=TRUE}
model
```
###Cross Validating the Model
```{r cross validation, cache = TRUE, eval = TRUE,echo=TRUE}
predictionCV<-predict(model,testing)
confusionMatrix(predictionCV,testing$classe)
```

###Predicting Results for the Quiz Data
```{r prediction }
prediction<-predict(model,testdata3)
```


## References
Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.
http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har
